### 2016-10-12 paper.tex
lost of changes

\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{width=7cm,compat=1.8}

@@

\newtheorem{cor}{Corallary}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Condition}

@@

%\newcommand{\smin}[1]{s_\text{min}({#1})}
\newcommand{\smin}{s_\text{min}}
\newcommand{\smax}{s_\text{max}}
\newcommand{\qhi}{q_\text{hi}}
\newcommand{\qlo}{q_\text{lo}}
\newcommand{\zero}{\text{\textbf{0}}}
\newcommand{\Zproj}{Z^{\textit{proj}}}
\newcommand{\Zreopt}{Z^{\textit{owa}}}
\newcommand{\nreopt}{n^{\textit{owa}}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\matW}{\hat W}
\newcommand{\matV}{\hat V}
\newcommand{\W}{{\hat \Theta^{\textit{owa}}}}
\newcommand{\Wowa}{{\hat \Theta^{\textit{owa}}}}
\newcommand{\Waff}{\mathcal{\hat A}}
\newcommand{\WaffE}{{\mathcal{\hat A}_\E}}
\newcommand{\Wave}{{\mathcal{\hat W}^{ave}}}
\newcommand{\Wtave}{{\mathcal{W}^{ave,*}}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\theta}
\newcommand{\wkl}{\hat\w^{kl}}
\newcommand{\ahat}{\hat\alpha}
\newcommand{\afull}{\ahat^{\textit{full}}}
\newcommand{\wowa}{\hat\w^{owa}}
\newcommand{\wowafull}{\hat\w^{\textit{owa,full}}}
\newcommand{\wowastar}{\hat\w^{\textit{owa,*}}}
\newcommand{\wave}{\hat\w^{ave}}
\newcommand{\wtave}{\E\hat\w^{ave}}
\newcommand{\waver}{\hat\w^{ave,r}}
\newcommand{\wboot}{\hat\w^{boot}}
\newcommand{\wmle}{\hat\w^{mle}}
\newcommand{\wmler}{\hat\w^{mle,r}}
\newcommand{\wstar}{{\w^{*}}}
\newcommand{\wq}{\hat\w^{q}}
\newcommand{\wqstar}{\hat\w^{q^*}}
\newcommand{\dist}{\mathcal{D}}

@@

\newcommand{\I}{\mathcal I}
%\newcommand{\I^{-1}}{\I^{-1}}
\newcommand{\law}{\ensuremath{\xrightarrow{L}}}
\newcommand{\normal}[2]{\ensuremath{\mathcal{N}\left({{#1}},{{#2}}\right)}}
\newcommand{\subnormal}[1]{\ensuremath{\subG\left({{#1}}\right)}}
\newcommand{\trans}[1]{\ensuremath{{#1}^{\mathsf{T}}}}
\newcommand{\ltwo}[1]{{\lVert {#1} \rVert}}
\newcommand{\ltwobig}[1]{{\left\lVert {#1} \right\rVert}}
\newcommand{\lone}[1]{{\lVert {#1} \rVert}_1}
\newcommand{\lzero}[1]{{\lVert {#1} \rVert}_0}
\newcommand{\proj}[1]{\pi_{{#1}}}
\newcommand{\prob}[1]{\Pr\left[{#1}\right]}

@@

\newcommand{\ignore}[1]{}
\newcommand{\fixme}[1]{\textbf{FIXME:} {#1}}

@@

@incollection{vershynin2010introduction,
  title={Introduction to the non-asymptotic analysis of random matrices},
  author={Vershynin, Roman},
  journal={arXiv preprint arXiv:1011.3027},
  chapter=5,
  booktitle={Compressed Sensing, Theory and Applications},
  editor={Y. Eldar and G. Kutyniok},
  year={2012}
}

@@

%\aistatstitle{Instructions for paper submissions to AISTATS 2015}
\aistatstitle{Communication-Efficient Distributed Maximum Likelihood Estimation with the Optimal Weigted Average}

@@

\begin{abstract}
We present a distributed algorithm for maximum likelihood estimation (MLE) with low communication complexity.
Our algorithm is an easy to implement modification of the standard averaging method.
In the averaging method, each machine in a cluster first solves the MLE problem on a subset of the data;
the resulting parameter vectors are then averaged together.
In our method, we perform a second round of optimization to find the \emph{optimal weighted average} of the parameter vectors.
This second optimization happens in a low dimensional space,
and so it adds negligible computational and communication costs.
%Given $m$ machines (each with $n$ data points of dimension $d$),
%we show that our algorithm reduces bias at a rate of $O(\frac{1-m/d}{n})$.
Whereas the averaging method reduces only the variance of the estimator,
our method reduces both the bias and the variance.
Our analysis is more general than the analysis of similar communication efficient algorithms that also reduce the bias.
Notably, we do not assume (as do all previous analyses) that the likelihood function is concave or that any quantities are bounded.
In practice,
this bias reduction makes our algorithm more robust to the choice of regularization parameter,
speeding up the model selection process.
\end{abstract}

@@

Many modern datasets are too large to fit in the memory of a single machine,
so they must be partitioned onto many machines.
%These datasets are typically partitioned onto many machines.
%Statistical learning algorithms are then run in a distributed manner.
%There are two main strategies for these distributed algorithms.
To analyze these datasets, we need distributed algorithms.
Existing distributed algorithms can be classified as either interactive or non-interactive.
In this paper we propose an algorithm that exhibits the benefits of both types.

\emph{Interactive algorithms} require many rounds of communication between machines.
These algorithms often resemble standard iterative algorithms where each iteration is followed by a communication step.
The appeal of interactive algorithms is that they enjoy the same statistical regret bounds as standard sequential algorithms.
Unfortunately, these algorithms can be too slow in practice because communication is the main bottleneck in modern distributed architectures,
and these algorithms require many rounds of communication.

\emph{Non-interactive} algorithms require only a single round of communication.
They are significantly faster than interactive algorithms
but have worse regret bounds.
Recent work (discussed in Section \ref{sec:bounds}) has shown that no non-interactive algorithm can achieve regret bounds comparable to an interactive one.
The popular large scale learning system Vowpal Wabbit combines both types of algorithms to achieve good real world performance.
A non-interactive algorithm is run first to provide a good warm start to a subsequent interactive algorithm \citep{vw,agarwal2014reliable}.
%Synonyms for non-interactive algorithms used in the literature include \emph{one-shot distributed learning} and \emph{communication efficient learning}.

In this paper, we propose a \emph{semi-interactive} distributed algorithm called \emph{optimal weighted averaging} (OWA).
Our algorithm performs two rounds of communication,
so it is not subject to the existing regret bounds of non-interactive algorithms.
The amount of information communicated in the second round is small, however,
so our algorithm retains the speed advantages of non-interactive algorithms.

In the next section, we formally describe our problem setting and proposed solution.
In Section 3, we compare our solution to related solutions.
We highlight how the analysis of existing algorithms requires more limiting assumptions than our own,
and show in detail why most existing non-interactive regret bounds do not apply to our problem.
Section 4.1 presents a novel, very general analysis of the averaging estimator.
Section 4.2 shows that two parameters to our algorithm $m$ and $\nreopt$ act as ``knobs'' that let the user easily trade off statistical performance for communication cost.
Section 5 shows experimentally that our algorithm performs well.

\section{Problem Statement and Solution}
%\section{Problem Setting and Our Solution}

@@

Let $\Y\subseteq\mathbb{R}$ be the space of response variables,
$X\subseteq\mathbb{R}^d$ be the space of the covariates,
and $\Theta\subseteq\mathbb{R}^d$ be the parameter space.
We assume a linear model.
The log-likelihood of data point $(\x,y)\in\X\times\Y$ given the model's true parameter $\wstar\in\Theta$ is denoted by $f(y,\trans\x\wstar)$.
Our analysis in Section \ref{sec:anal} places very mild restrictions on $f$.
In particular, $f$ need not be concave with respect to $\w$.
%For our experiments in Section \ref{sec:exp}, however, we restrict ourselves to the likelihood for logistic regression
%\begin{equation}
%f(y,\trans\x\w) =
%\end{equation}
Let $Z\subset\X\times\Y$ be a dataset of $mn$ i.i.d. observations.
Finally, let $R : \Theta \to \mathbb{R}$ be a regularization function (typically the L1 or L2 norm)
and $\lambda\in\mathbb{R}$ be the regularization strength.
Then the maximum likelihood estimator is
\begin{equation}
%\wmle=\argmax_\w \sum_{i=1}^{nm} g(y_i-\trans\x_i\w)
\wmle=\argmax_\w \sum_{(\x,y)\in Z} f(y,\trans\x\w)
+ \lambda R(\theta)
.
\end{equation}
Assume that $Z$ has been partitioned onto $m$ machines so that each machine $i$ has dataset $Z_i$ of size $n$, and all the $Z_i$ are disjoint.
Then each machine calculates the local estimator
\begin{equation}
\wmle_i = \argmax_\w \sum_{(\x,y) \in Z_i} f(y,\trans\x\w)
+ \lambda R(\theta)
.
\end{equation}
Solving for $\wmle_i$ requires no communication with other machines.
%Our goal is to provide a function $q : \Theta^m \to \Theta$ that ``merges'' these estimates.
%The resulting merged estimate should have error as close to $\wmle$ as possible.
%A non-interactive distributed estimator merges these $\wmle_i$ into a single improved estimate.
Our goal is to merge the $\wmle_i$ into a single improved estimate.
The baseline example is the averaging estimator
\begin{equation}
\wave = \frac{1}{m}\sum_{i=1}^m \wmle_i
.
\end{equation}
This estimator is well studied, and in Section \ref{sec:anal} we compare this previous work to our own.
Here we briefly recall that the analysis of an estimator's error $\ltwo{\wstar-\wave}$ is typically decomposed into bias $\ltwo{\wstar - \E\wave}$ and variance $\ltwo{\E\wave-\wave}$ components.
The $\wave$ estimator is known to reduce the variance but not the bias.
Our goal is to design an estimator that reduces both variance and bias.

@@

To motivate our estimator,
we first describe a simpler estimator that uses the entire dataset for the second round of optimization.
%The idea of our proposed solution is to find the best linear combination of the $\wmle_i$s using a second round of optimization.
Define the matrix $\matW : \mathbb{R}^{d\times m}$ to have $i$th column equal to $\wmle_i$.
Now consider the estimator
\begin{equation}
\wowafull = \matW \afull
,
\end{equation}
where
\begin{equation}
\label{eq:afull}
\afull = \argmax_\alpha \sum _{(\x,y)\in Z} f\left(y,\trans\x \matW \alpha \right)
+
\lambda R(\matW\alpha)
.
\end{equation}
%This is the maximum likelihood estimator when we restrict our
Notice also that $\wowafull$ is the maximum likelihood estimator when the parameter space $\Theta$ is restricted to $\W = \vecspan \{\wmle_i\}_{i=1}^m$.
In other words, the $\afull$ vector contains the optimal weights to apply to each $\wmle_i$ when averaging.
Figure \ref{fig:contour} shows graphically that no other estimator in $\W$ can have higher regularized empirical likelihood than $\wowafull$.

\begin{figure}
\hspace{-0.1in}
\begin{tikzpicture}
    [
    dot/.style = {minimum width=0.15cm,inner sep=0pt,line width=0pt,fill,circle,black,font=\small}
    ]
\draw[gray] plot [smooth cycle,tension=1] coordinates {(-0.5,-0.75) (-1.05,1) (-0.1,-0.1) (0.75,-0.5) (0.45,-0.7) };
\draw[gray] plot [smooth cycle,tension=1] coordinates {(-0.85,-0.85) (-1.3,1.55) (0.25,0) (2.5,-0.5) (1,-0.95)};
\draw[gray] plot [smooth cycle,tension=1] coordinates {(-1.15,-1.2) (-1.5,2) (1,0) (5,0) (3,-0.95)};

@@

%\node[dot] (wavestar) at (4,0.5) {};
%\node                 at (4.5,0.3) {$\E\wave$};
\node[dot] (wave) at (3.8,1.25) {};
\node at (4.0,1.6) {$\wave$};
\node[dot] (wave) at (2.80,1.40) {};
\node at (3.1,1.75) {$\wmle_1$};
\node[dot] (wave) at (4.75,1.10) {};
\node at (4.9,1.45) {$\wmle_2$};
%\node[dot] (wave) at (3.8,1.25) {};

\node[dot] at (-1.35,2.03) {};
\node at (-0.9,2.4) {$\wowafull$};

\node[dot] at (-0.55,1.9) {};
\node at (-0.1,2.1) {$\wowa$};

\node at (4,-1.3) {$f(y,\trans\x\w)+\lambda R(\theta)$};
\node at (5.65,1.0) {$\W$};

\end{tikzpicture}
\caption{
    Our method performs a second round of optimization to find the best parameter vector in $\W$.
    Since $\wave\in\W$, the estimator $\wowafull$ has higher empirical likelihood than $\wave$.
    Since $\W$ has low dimension, we can use relatively few data points in the second round of optimization to ensure that with high probability $\wowa$ has higher empirical likelihood than $\wave$.
}
\label{fig:contour}
\end{figure}

Calculating the weights $\afull$ directly is infeasible because it requires access to the full dataset.
Fortunately, we don't need to consider all the data points for an accurate estimator.
The parameter space $\W$ is $m$-dimensional.
So intuitively, we only need $O(m)$ data points to solve the second optimization to our desired accuracy.
%(This intuition is formalized in Section \ref{sec:anal}.)
This intuition motivates the following estimator.
Let $\Zreopt_i\subset Z_i$ be a set of $\nreopt$ data points uniformly sampled from $Z_i$ without replacement,
and let $\Zreopt$ be the union of the $\Zreopt_i$s.
Then the OWA estimator is defined as
\begin{equation}
\wowa = \matW \ahat,
\end{equation}
where
\begin{equation}
\label{eq:ahat}
\ahat = \argmax_\alpha \sum _{(\x,y)\in \Zreopt} f\left(y,\trans\x \matW \alpha \right)
+ \lambda R (\matW\alpha)
.
\end{equation}

Algorithm \ref{fig:alg2} shows the steps for efficiently computing $\wowa$.
In the first round, each machine calculates $\wmle_i$ independently and broadcasts the result to every other machine.
Since the parameter vector has $d$ dimensions and there are $m$ machines, a total of $O(md)$ bits are transmitted.
In the second round, each machine projects its local dataset $\Zreopt_i$ onto the space $\W$.
These projected data points are then transmitted to a predesignated master machine.
The projected data points each have dimension $m$, so a total of $O(m^2\nreopt)$ bits are transmitted.
The master machine now has all of the information to complete the optimization.

\begin{algorithm}[t]
\caption{Distributed calculation of $\wowa$}
\label{alg:distributed}
\begin{algorithmic}
\State Each machine $i$ independently
\State ~~~~~loads dataset $Z_i$
\State ~~~~~calculates $\wmle_i$
\State All machines broadcast $\wmle_i$ to all other machines
\State Each machine $i$ independently
\State ~~~~~constructs $\matW$
\State ~~~~~randomly selects a dataset $Z^{reopt}_i\subset Z_i$
\State ~~~~~calculates $\Zproj_i=\{(\trans\x\matW,y) : (\x,y)\in\Zreopt\}$
\State All machines broadcast $\Zproj_i$ to a master machine
\State The master calculates $\wowa$
\end{algorithmic}
\label{fig:alg2}
\end{algorithm}

\subsection{Implementation Tips}

Equations \ref{eq:afull} and \ref{eq:ahat} cannot be solved directly using off the shelf optimizers because existing optimizers do not support the non-standard regularization term $R(\matW\alpha)$.
In practice, it is sufficient to approximate this regularization by L2 regularization directly on the $\alpha$ vector:
\begin{equation}
\lambda R(\matW\alpha) \approx \lambda_2 \ltwo{\alpha}
.
\end{equation}
Intuitively, this is because even when we want the parameter vector $\w$ to be sparse (and so are regularizing by $R = $ the L1 norm), we have no reason to believe that the $\alpha$ vector should be sparse.
The desired sparsity is induced by the regularization when solving for $\wmle_i$s and maintained in any linear combination of the $\wmle_i$s.
The new $\lambda_2$ regularization parameter should be set by cross validation.
This will be a fast procedure, however, because there are few data points to optimize over,
and the L2 regularized problem is much easier to solve than the L1 problem.
The new $\lambda_2$ regularization parameter should be set by cross validation.
This will be a fast procedure, however, because there are few data points to optimize,
and the L2 regularized problem is much easier to solve than the L1 problem.
With this minor modification, our distributed estimator can be implemented using any existing optimizer.

It is also important when performing the second round of optimization to not extend the parameter vector with an intercept term as is typically done.
It is fine to do this for the first round as long as the intercepts also get scaled.

\section{RELATED WORK}

>>>>

\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{width=7cm,compat=1.8}
\definecolor{darkgreen}{RGB}{0,127,0}

@@

\newtheorem{cor}{Corallary}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Condition}
\newtheorem{defn}{Definition}

@@

%\newcommand{\smin}[1]{s_\text{min}({#1})}
\newcommand{\smin}{s_\text{min}}
\newcommand{\smax}{s_\text{max}}
\newcommand{\qhi}{q_\text{hi}}
\newcommand{\qlo}{q_\text{lo}}
\newcommand{\zero}{\text{\textbf{0}}}
\newcommand{\Zproj}{Z^{\textit{proj}}}
\newcommand{\Zowa}{Z^{\textit{owa}}}
\newcommand{\nowa}{n^{\textit{owa}}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\matW}{\hat W}
\newcommand{\matV}{\hat V}
\newcommand{\W}{{\hat \Theta^{\textit{owa}}}}
\newcommand{\Wowa}{{\hat \Theta^{\textit{owa}}}}
\newcommand{\Waff}{\mathcal{\hat A}}
\newcommand{\WaffE}{{\mathcal{\hat A}_\E}}
\newcommand{\Wave}{{\mathcal{\hat W}^{ave}}}
\newcommand{\Wtave}{{\mathcal{W}^{ave,*}}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\theta}
\newcommand{\wkl}{\hat\w^{kl}}
\newcommand{\ahat}{\hat\alpha}
\newcommand{\astar}{\alpha^*}
\newcommand{\afull}{\ahat^{\textit{full}}}
\newcommand{\wowa}{\hat\w^{owa}}
\newcommand{\wowafull}{\hat\w^{\textit{owa,full}}}
\newcommand{\wowastar}{\hat\w^{\textit{owa,*}}}
\newcommand{\wave}{\hat\w^{ave}}
\newcommand{\wtave}{\E\hat\w^{ave}}
\newcommand{\waver}{\hat\w^{ave,r}}
\newcommand{\wboot}{\hat\w^{boot}}
\newcommand{\wmle}{\hat\w^{mle}}
\newcommand{\wmler}{\hat\w^{mle,r}}
\newcommand{\wstar}{{\w^{*}}}
\newcommand{\wq}{\hat\w^{q}}
\newcommand{\wqstar}{\hat\w^{q^*}}
\newcommand{\dist}{\mathcal{D}}

@@

\newcommand{\I}{\mathcal I}
%\newcommand{\I^{-1}}{\I^{-1}}
\newcommand{\law}{\ensuremath{\xrightarrow{L}}}
\newcommand{\normal}[2]{\ensuremath{\mathcal{N}\left({{#1}},{{#2}}\right)}}
\newcommand{\subnormal}[1]{\ensuremath{\subG\left({{#1}}\right)}}
\newcommand{\trans}[1]{\ensuremath{{#1}^{\mathsf{T}}}}
\newcommand{\pinv}[1]{\ensuremath{{#1}^{\mathsf{\dagger}}}}
\newcommand{\ltwo}[1]{{\lVert {#1} \rVert}}
\newcommand{\ltwobig}[1]{{\left\lVert {#1} \right\rVert}}
\newcommand{\lone}[1]{{\lVert {#1} \rVert}_1}
\newcommand{\lzero}[1]{{\lVert {#1} \rVert}_0}
\newcommand{\proj}[1]{\pi_{{#1}}}
\newcommand{\prob}[1]{\Pr\left[{#1}\right]}

@@

%\newcommand{\plots}[1]{}
\newcommand{\plots}[1]{#1}
\newcommand{\ignore}[1]{}
\newcommand{\fixme}[1]{\textbf{FIXME:} {#1}}

@@

@incollection{vershynin2010introduction,
  title={Introduction to the non-asymptotic analysis of random matrices},
  author={Vershynin, Roman},
  chapter=5,
  booktitle={Compressed Sensing, Theory and Applications},
  editor={Y. Eldar and G. Kutyniok},
  year={2012}
}

@@

%\aistatstitle{Instructions for paper submissions to AISTATS 2015}
\aistatstitle{Communication-Efficient Distributed Maximum Likelihood Estimation with the Optimal Weighted Average}

@@

\begin{abstract}
We present a distributed algorithm for maximum likelihood estimation (MLE) with low communication cost. % called optimal weighted averaging (OWA).
In our algorithm, each machine in a cluster first independently solves the MLE problem on a subset of the data;
a second round of optimization then finds the optimal weighted average of the parameter vectors.
%In the first step, each machine independently solves the MLE problem on a local subset of the data.
%Then a second round of optimization finds the optimal weighted average of the parameter vectors from step 1.
%The resulting parameter vectors are then merged together
%OWA proceeds in two rounds.
%In the first round, each machine solves the MLE problem on a local subset of the data.
%In the second round, each machine solves a modified MLE problem.
%The number of machines $m$
%We then perform a second round of optimization to find the optimal weighted average of the results.
%Our algorithm is an easy to implement improvement to the standard averaging method.
%In the averaging method, each machine in a cluster first solves the MLE problem on a subset of the data;
%the resulting parameter vectors are then averaged together.
%In our method, we perform a second round of optimization to find the \emph{optimal weighted average} of the parameter vectors.
%This second optimization happens in a low dimensional space,
%and so it adds negligible computational and communication costs.
%Whereas the averaging method reduces only the variance of the estimator,
%our method reduces both the bias and the variance.
%A major advantage of OWA is that two tunable parameters let the user trade better statistical p
We show analytically that this method reduces both the bias and variance of the estimator when compared to the estimator trained by a single machine.
(The straight average is known to reduce only the variance.)
Our analysis relies on simple properties of sub-Gaussian random vectors.
It is therefore simpler and more general than the analysis of similar distributed algorithms.
Notably, we do not assume (as do all previous analyses) that the likelihood function is concave or that any quantities are bounded.
A major practical advantage of our method is that it is robust to the amount of regularization,
which speeds up model selection.
\end{abstract}

@@

Many modern datasets are too large to fit in the memory of a single machine,
so they must be partitioned onto many machines.
%These datasets are typically partitioned onto many machines.
%Statistical learning algorithms are then run in a distributed manner.
%There are two main strategies for these distributed algorithms.
To analyze these datasets, we need distributed algorithms.
Existing distributed algorithms can be classified as either interactive or non-interactive depending on their communication complexity.
In this paper we propose an algorithm that exhibits the benefits of both types.

\emph{Interactive} algorithms require many rounds of communication between machines.
These algorithms often resemble standard iterative algorithms where each iteration is followed by a communication step.
The appeal of interactive algorithms is that they enjoy the same statistical regret bounds as standard sequential algorithms.
But there are two downsides.
First, these algorithms can be too slow in practice because communication is the main bottleneck in modern distributed architectures.
Second, these algorithms require special implementations and do not work with off-the-shelf statistics libraries provided by (for example) Python, R, and Matlab.

\emph{Non-interactive} algorithms require only a single round of communication.
They are significantly faster than interactive algorithms and  easily implemented with standard libraries.
The downside is worse regret bounds.
Recent work (discussed in Section \ref{sec:bounds}) has shown that no non-interactive algorithm can achieve regret bounds comparable to an interactive one.

In this paper, we propose a \emph{semi-interactive} distributed algorithm called \emph{optimal weighted averaging} (OWA).
Our algorithm performs two rounds of communication,
so it is not subject to the existing regret bounds of non-interactive algorithms.
The algorithm has two tunable parameters that let the user trade better statistical performance for worse communication complexity.
%The amount of information communicated in the second round is small, however,
%so our algorithm retains the speed advantages of non-interactive algorithms.
The OWA algorithm is easily implemented in a MapReduce architecture with standard packages.

%The popular large scale learning system Vowpal Wabbit combines both types of algorithms to achieve good real world performance.
%A non-interactive algorithm is run first to provide a good warm start to a subsequent interactive algorithm \citep{vw,agarwal2014reliable}.

In the next section, we formally describe our OWA algorithm.
In Section 3, we compare OWA to existing distributed algorithms.
We highlight how the analysis of existing algorithms requires more limiting assumptions than our own,
and show in detail why existing non-interactive regret bounds do not apply to OWA.
%Section 4.1 presents a novel, more general analysis of the widely used averaging estimator.
%Section 4.2 shows that OWA's regret bounds interpolates between the averaging estimator's regret and the optimal regret.
Section 4 shows that OWA's regret bounds interpolate between the averaging estimator's regret and the optimal regret.
As part of the analysis, we provide novel, more general regret bounds for the averaging estimator.
Section 5 shows experimentally that our algorithm performs well.
We emphasize that our algorithm is robust to the strength of regularization,
which is one of the reasons it performs well in practice.

\section{THE OWA ALGORITHM}
%\section{Problem Setting and Our Solution}

@@

This section first formally introduces the problem of communication efficient distributed estimation,
then describes our proposed OWA distributed estimator.

\subsection{Problem Setting}

Let $\Y\subseteq\mathbb{R}$ be the space of response variables,
$X\subseteq\mathbb{R}^d$ be the space of covariates,
and $\Theta\subseteq\mathbb{R}^d$ be the parameter space.
We assume a linear model.
The log-likelihood of data point $(\x,y)\in\X\times\Y$ given the model's true parameter $\wstar\in\Theta$ is denoted by $f(y,\trans\x\wstar)$.
Our analysis in Section \ref{sec:anal} places very mild restrictions on $f$.
In particular, $f$ need not be concave with respect to $\w$.
%For our experiments in Section \ref{sec:exp}, however, we restrict ourselves to the likelihood for logistic regression
%\begin{equation}
%f(y,\trans\x\w) =
%\end{equation}
Let $Z\subset\X\times\Y$ be a dataset of $mn$ i.i.d. observations.
Finally, let $R : \Theta \to \mathbb{R}$ be a regularization function (typically the L1 or L2 norm)
and $\lambda\in\mathbb{R}$ be the regularization strength.
Then the regularized maximum likelihood estimator (MLE) is
\begin{equation}
%\wmle=\argmax_\w \sum_{i=1}^{nm} g(y_i-\trans\x_i\w)
\wmle=\argmax_\w \sum_{(\x,y)\in Z} f(y,\trans\x\w)
+ \lambda R(\theta)
.
\end{equation}
In the remainder of this paper, it should be understood that all MLEs are regularized.

Assume that $Z$ has been partitioned onto $m$ machines so that each machine $i$ has dataset $Z_i$ of size $n$, and all the $Z_i$ are disjoint.
Then each machine calculates the local MLE
\begin{equation}
\wmle_i = \argmax_\w \sum_{(\x,y) \in Z_i} f(y,\trans\x\w)
+ \lambda R(\theta)
.
\end{equation}
Solving for $\wmle_i$ requires no communication with other machines.
%Our goal is to provide a function $q : \Theta^m \to \Theta$ that ``merges'' these estimates.
%The resulting merged estimate should have error as close to $\wmle$ as possible.
%A non-interactive distributed estimator merges these $\wmle_i$ into a single improved estimate.
Our goal is to merge the $\wmle_i$ into a single improved estimate.
%A non-interactive distributed estimator depends on the $\wmle_i$s and nothing else.
A baseline merging procedure is the averaging estimator
\begin{equation}
\wave = \frac{1}{m}\sum_{i=1}^m \wmle_i
.
\end{equation}
This estimator is well studied, and in Section \ref{sec:relwork} we compare this previous work to our own.
Here we briefly recall that the analysis of an estimator's error $\ltwo{\wstar-\wave}$ is typically decomposed into bias $\ltwo{\wstar - \E\wave}$ and variance $\ltwo{\E\wave-\wave}$ components.
The $\wave$ estimator is known to reduce the variance but not the bias.
Our goal is to design an estimator that reduces both variance and bias.

@@

To motivate our estimator,
we first present an estimator that uses the entire dataset for the second round of optimization.
%The idea of our proposed solution is to find the best linear combination of the $\wmle_i$s using a second round of optimization.
Define the matrix $\matW : \mathbb{R}^{d\times m}$ to have $i$th column equal to $\wmle_i$.
Now consider the estimator
\begin{equation}
\wowafull = \matW \afull
,
\end{equation}
where
\begin{equation}
\label{eq:afull}
\afull = \argmax_\alpha \sum _{(\x,y)\in Z} f\left(y,\trans\x \matW \alpha \right)
+
\lambda R(\matW\alpha)
.
\end{equation}
Notice that $\wowafull$ is the maximum likelihood estimator when the parameter space $\Theta$ is restricted to $\W = \vecspan \{\wmle_i\}_{i=1}^m$.
In other words, the $\afull$ vector contains the optimal weights to apply to each $\wmle_i$ when averaging.
Figure \ref{fig:contour} shows graphically that no other estimator in $\W$ can have higher regularized empirical likelihood than $\wowafull$.

\begin{figure}
\hspace{-0.1in}
\begin{tikzpicture}
    [
    dot/.style = {minimum width=0.15cm,inner sep=0pt,line width=0pt,fill,circle,black,font=\small}
    , yscale=0.65
    ]
\small
\draw[gray] plot [smooth cycle,tension=1] coordinates {(-0.5,-0.75) (-1.05,1) (-0.1,-0.1) (0.75,-0.5) (0.45,-0.7) };
\draw[gray] plot [smooth cycle,tension=1] coordinates {(-0.85,-0.85) (-1.3,1.55) (0.25,0) (2.5,-0.5) (1,-0.95)};
\draw[gray] plot [smooth cycle,tension=1] coordinates {(-1.15,-1.2) (-1.5,2) (1,0) (5,0) (3,-0.95)};

@@

%\node[dot] (wavestar) at (4,0.5) {};
%\node                 at (4.5,0.3) {$\E\wave$};
\node[dot] (wave) at (3.8,1.25) {};
\node at (4.0,1.7) {$\wave$};
\node[dot] (wave) at (2.80,1.40) {};
\node at (3.1,1.85) {$\wmle_1$};
\node[dot] (wave) at (4.75,1.10) {};
\node at (4.9,1.55) {$\wmle_2$};
%\node[dot] (wave) at (3.8,1.25) {};

\node[dot] at (-1.35,2.03) {};
\node at (-1.0,2.5) {$\wowafull$};

\node[dot] at (-0.55,1.9) {};
\node at (-0.1,2.3) {$\wowa$};

\node at (4,-1.3) {$f(y,\trans\x\w)+\lambda R(\theta)$};
\node at (5.65,1.0) {$\W$};
\end{tikzpicture}
\vspace{-0.35in}
\caption{
    Our method performs a second round of optimization to find the best parameter vector in $\W$.
    %Since $\wave\in\W$, the estimator $\wowafull$ has higher empirical likelihood than $\wave$.
    Since $\W$ has low dimension, we can use relatively few data points in the second round of optimization to ensure that with high probability $\wowa$ has higher empirical likelihood than $\wave$.
    %Since $\W$ has low dimension, we can use relatively few data points in the second round of optimization to ensure that $\ltwo{\wowafull-\wowa}$ is small.
}
\label{fig:contour}
\end{figure}

Calculating the weights $\afull$ directly is infeasible because it requires access to the full dataset.
Fortunately, we don't need to consider all the data points for an accurate estimator.
The parameter space $\W$ is $m$-dimensional.
So intuitively, we only need $O(m)$ data points to solve the second optimization to our desired accuracy.
%(This intuition is formalized in Section \ref{sec:anal}.)
This intuition motivates the OWA estimator.
Let $\Zowa_i\subset Z_i$ be a set of $\nowa$ data points uniformly sampled from $Z_i$ without replacement,
and let $\Zowa$ be the union of the $\Zowa_i$s.
Then the OWA estimator is defined as
\begin{equation}
\wowa = \matW \ahat,
\end{equation}
where
\begin{equation}
\label{eq:ahat}
\ahat = \argmax_\alpha \sum _{(\x,y)\in \Zowa} f\left(y,\trans\x \matW \alpha \right)
+ \lambda R (\matW\alpha)
.
\end{equation}

Algorithm \ref{fig:alg2} shows the steps for efficiently computing $\wowa$.
In the first round, each machine calculates $\wmle_i$ independently and broadcasts the result to every other machine.
Since the parameter vector has $d$ dimensions and there are $m$ machines, a total of $O(md)$ bits are transmitted.
In the second round, each machine projects its local dataset $\Zowa_i$ onto the space $\W$.
These projected data points are then transmitted to a predesignated master machine.
The projected data points each have dimension $m$, so a total of $O(m^2\nowa)$ bits are transmitted.
The master machine now has all of the information to complete the optimization.
In total, $O(md + m^2\nowa)$ bits were transmitted.
When $d<m\nowa$, this is the same order of bits as the averaging estimator.

\begin{algorithm}[t]
\caption{Distributed calculation of $\wowa$}
\label{alg:distributed}
\begin{algorithmic}
\State Round 1, each machine $i$ independently:
\State ~~~~~loads dataset $Z_i$
\State ~~~~~calculates $\wmle_i$
\State ~~~~~broadcasts $\wmle_i$ to all other machines
\State Round 2, each machine $i$ independently:
\State ~~~~~constructs $\matW$
\State ~~~~~randomly selects a dataset $\Zowa_i\subset Z_i$
\State ~~~~~calculates $\Zproj_i=\{(\trans\x\matW,y) : (\x,y)\in\Zowa\}$
\State ~~~~~broadcasts $\Zproj_i$ to a master machine
\State The master calculates $\wowa$
\end{algorithmic}
\label{fig:alg2}
\end{algorithm}

%\subsection{Implementation Tips}
%
Equations \ref{eq:afull} and \ref{eq:ahat} cannot be solved directly using off the shelf optimizers because existing optimizers do not support the non-standard regularization term $R(\matW\alpha)$.
In practice, it is sufficient to approximate this regularization by L2 regularization directly on the $\alpha$ vector:
\begin{equation}
\lambda R(\matW\alpha) \approx \lambda_2 \ltwo{\alpha}
.
\label{eq:approxreg}
\end{equation}
Intuitively, this is because even when we want the parameter vector $\w$ to be sparse (and so are regularizing by $R = $ the L1 norm), we have no reason to believe that the $\alpha$ vector should be sparse.
The desired sparsity is induced by the regularization when solving for $\wmle_i$s and maintained in any linear combination of the $\wmle_i$s.
The new $\lambda_2$ regularization parameter should be set by cross validation.
This will be a fast procedure, however, because there are few data points to optimize over,
and the L2 regularized problem is much easier to solve than the L1 problem.
With this minor modification, our distributed estimator can be implemented using any existing optimizer.
%Our experimental results in Section \ref{sec:exp} use this modified version of $\hat\alpha$.
